---
title: "PCAwR"
author: "Laci Cartmell"
date: "10/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Presentation Rubric
Each student will develop a lesson and teach the class about an R package, specialized analysis, or data visualization approach that they have or will apply to their own research.

### Your lesson should include: 

1. An introductory lecture/walk through of the tool, analyses, or visualization.
- Aim for ~40 minute presentation to allow time for other lessons and for folks to work through exercises at the end of the session.
- Detailed lecture notes.
written in Rmarkdown
posted to you Github webpage in html format.
2. An exercise where folks can practice implementing the to tool, analyses, or visualization.
written in Rmarkdown
posted to you Github webpage in html format.
Where appropriate provide a sample data set (ideally real data).
- Provide assistance to students to help them complete the exercise
- Please select your topic and confirm it with me at least one month before your scheduled presentation.

##  Grading
  Item	Points
  Lecture presentation	40
  Lecture notes	40
  Exercise	20
  Total	100

##PCA 

Principal Components Analysis (PCA) visualizes large datasets with complex interactions

It finds the lowest variation and 

 embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots
embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
